---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
library(lubridate)
dat <- read.table(unz("activity.zip", "activity.csv"), sep=",", header=TRUE, stringsAsFactors = FALSE)
dat$date <- ymd(dat$date)
head(dat)
```

## What is mean total number of steps taken per day?
```{r}
totalPerDay <- tapply(dat$steps, factor(dat$date), sum, na.rm = TRUE)
meanPerDay <- mean(totalPerDay)
medianPerDay <- median(totalPerDay)
hist(totalPerDay, main="Histogram for Total Steps Per Day", xlab="Number of Steps Taken", breaks = 10)

```

The mean total steps per day is `r meanPerDay` and the median total steps per day is `r medianPerDay`.

## What is the average daily activity pattern?
```{r} 
totalsPerInterval <- tapply(dat$steps, factor(dat$interval), mean, na.rm=TRUE)
plot(as.numeric(names(totalsPerInterval)), 
     totalsPerInterval, 
     type="l",
     main = "Mean Steps by 5 Minute Interval",
     xlab = "Interval (i.e. minutes after midnight, in five-minute groups)",
     ylab = "Mean Steps During Interval")
intervalWithMostSteps <- names(totalsPerInterval)[which.max(totalsPerInterval)]

```
The interval with the highest mean number of steps is `r intervalWithMostSteps`

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
